Fast Bayesian Inference for Exoplanet Discovery in Radial Velocity Data

### 2014-11-14 Paper/ms.tex
Working...

% Bold symbols
\renewcommand{\btheta}{\boldsymbol{\theta}}
\newcommand{\bdata}{\boldsymbol{D}}
\newcommand{\bx}{\{x_i\}}

@@

\section{Inference}
Bayesian inference is the use of probability theory to describe uncertainty
\citep{sivia, ohagan}. In this framework, we approach data analysis problems by first
constructing a {\it hypothesis space}, which is the set of possible answers
to the problem we are considering. Normally, this is the set of possible
values of a vector of parameters $\btheta$ whose values we want
to know. We then assign probability distributions
called the {\it prior} and the {\it sampling distribution}. The prior
distribution $p(\btheta)$ describes
our initial uncertainty about which values of the parameters
$\btheta$ are plausible, and the
sampling distribution $p(\bdata | \btheta)$ describes
our initial uncertainty about the data set we're going to observe, as a
function of the unknown parameters $\btheta$.
When the data is known, our state of knowledge about the parameter is
updated from the prior $p(\theta)$ to the posterior distribution given by
Bayes' rule:
\begin{eqnarray}
p(\btheta | \bdata) &=&
\frac{p(\btheta)p(\bdata | \btheta)}
{p(\bdata)}
\end{eqnarray}
where $p(\bdata | \btheta)$ is called the likelihood, once the actual dataset
is substituted in.
The denominator, often called the {\it evidence} or
{\it marginal likelihood}, is given by the expected value of the likelihood
with respect to the prior:
\begin{eqnarray}
\mathcal{Z} = p(\bdata) = \int p(\btheta) p(\bdata | \btheta) d^n \btheta
\end{eqnarray}
where the integral is over the entire $n$-dimensional parameter space.

\subsection{Inferring the number of planets}
The number of orbiting planets, $N$, is an important parameter.
To calculate the posterior distribution for $N$, most authors
consider various trial values of $N$, and calculate the marginal likelihood
\begin{eqnarray}
p(\bdata | N) &=& \int p(\btheta | N) p(\bdata | \btheta, N) \, d^n \btheta
\end{eqnarray}
for each possible value of $N$
\citep[e.g.][]{2011MNRAS.415.2523G, 2011MNRAS.415.3462F, 2014MNRAS.437.3540F, fengji}, marginalising
over all other model parameters.
The posterior distribution for $N$ can then be found straightforwardly by
using Bayes' rule with $N$ as the only unknown parameter:
\begin{eqnarray}
p(N | \bdata) &=& \frac{p(N)p(\bdata | N)}{\sum_N p(N)p(\bdata | N)}.
\end{eqnarray}
Popular
methods for calculating the marginal likelihood are Nested Sampling
\citep{skilling} and ideas related to thermodynamic integration
\citep[e.g.][]{neal}.

@@

Imagine a high-dimensional unimodal posterior distribution that is composed of a
broad, high volume but low density ``slab'' with a narrow, low volume but high
density ``spike'' on top of it.
An example is a mixture of two concentric high-dimensional gaussians with
different widths. If you ran MCMC on such a posterior, it would be difficult
to jump between the slab and the spike components. If the MCMC is currently
in the spike region (or phase) it will be unable to escape: a proposed move
into the slab will be rejected because of the ratio of densities. Conversely,
if the MCMC was in the slab region, it would be unlikely to go into the spike
region, because its volume is so small: it would be very unlikely to
{\it propose} to move into the spike.

@@

\begin{table*}
\begin{tabular}{|l|l|l|}
\hline
Quantity	&	Meaning		& Prior\\
\hline
{\bf Hyperparameters}	&	\\
$N$		& Number of planets	& Uniform$(\{0, 1, ..., N_{\rm max}\})$\\
$\mu_T$		&	Median orbital period	& $\ln(\mu_T) \sim$ Cauchy$(5.9, 1)T(-15.3, 27.1)$\\
$w_T$		&	Diversity of orbital periods & $w_T \sim$ Uniform$(0.1, 3)$\\
$\mu_A$		&	Mean amplitude	& $\ln(\mu_A) \sim$ Cauchy$(0, 1)T(-21.2, 21.2)$\\
\hline
{\bf Planet Parameters}\\
$T_i$		&	Orbital period	&	$\ln(T_i) \sim $ Biexponential($\ln(\mu_T), w_T$)\\
$A_i$		&	Amplitude of signal	& Exponential$(\mu_A)$\\
$\phi_i$	&	Phase of signal	&	Uniform($0, 2\pi$)\\
$v_i$		&	Initial speed of orbit	&	Uniform$(0.4, 1)$\\
$\theta_i$	&	Viewing angle	&	Uniform$(0, 2\pi)$\\
\hline
{\bf Other}\\
$\sigma_{\rm extra}$	& ``Extra noise'' parameter	& $\ln(\sigma) \sim$ Cauchy$(0, 1)T(-21.2, 21.2)$\\
$\nu$		& Shape parameter for $t$-distribution for noise & $\ln(\nu) \sim$ Uniform$(\ln(0.01), \ln(1000))$\\
\hline
{\bf Data}\\
$Y_i$		& Radial velocity measurements	&
		Student-$t\left(m(t_i), \sqrt{\sigma_i^2 + \sigma_{\rm extra}^2}, \nu\right)$
\end{tabular}
\caption{All of the prior distributions in our Bayesian model.
The priors for the planet parameters are defined conditional on the values
of the hyperparameters. Uniform priors were used for parameters like phases.
For parameters where a rough initial guess is possible, heavy-tailed Cauchy
distributions were used so this information could be taken into account
in a non-dogmatic way. Time units are in days and amplitude units are
in metres per second. The maximum number of planets, $N_{\rm max}$, was
set to 15.\label{tab:priors}}
\end{table*}

@@

The posterior distribution for the logarithms of the periods is shown in
Figure~\ref{fig:nu_oph_periods}. Because of the label-switching degeneracy,
the posterior samples for all periods were combined to make this figure, which
shows three prominent peaks. Defining the log-periods by
$S_i =  \log_{10}\left[T/(\textnormal{1 day})\right]$,
Figure~\ref{fig:nu_oph_periods} is a Monte Carlo representation
of the mixture distribution
\begin{eqnarray}
f(S) &=& \sum_{N=0}^{10} p(N | \bdata)\sum_{i=1}^N p(S_i | N, \bdata).
\end{eqnarray}
The two rightmost peaks are the commonly accepted periods, and the third
prominent peak corresponds to a 75.7 $\pm$ 1.8 day period. The probability
this signal exists (technically, the probability a period exists between
67.6 and 85.1 days)
is 70\%, and given that it exists, its velocity amplitude
is 18.6 $\pm$ 8.3 m s$^{-1}$ and its eccentricity (or velocity parameter $v$)
is not well determined.

@@

Figure~\ref{fig:logl0} shows the relationship between the likelihood $L$
and the enclosed prior mass $X$ for the $\nu$ Oph analysis. These plots are
a standard output of Nested Sampling \citep{skilling} analyses, and provide
insights into the structure of the problem. Concave-up regions of this curve
indicate phase transitions which can cause severe problems for annealing-based
methods, and sometimes even for sampling the posterior distribution. In this
analysis, the models with a third signal exist to the left of the phase
transition at $\log(X) \approx -75$, and models without the signal exist to the
right of the phase transition. Mixing between these two phases is crucial for
accurately computing the posterior probability that the third signal exists.
However, if we had simply tried to sample the posterior distribution, it would
not mix between these two phases because the posterior is composed of a
large-volume low-likelihood ``slab'' and a low-volume high-likelihood ``spike''.
It is difficult for standard MCMC methods to escape the spike because of the
likelihood contrast, and it is also difficult to escape the slab and enter the
spike because of its small volume. Thus, even unimodal posterior distributions
can cause mixing problems analogous to multimodal situations if a phase
transition is present.

The marginal likelihood for our model was
$\ln(\mathcal{Z}) \approx -217.04$ and the information
(Kullback-Leibler divergence from the prior to the posterior)
was $\mathcal{H} \approx 80.0$ nats.

@@

The posterior distribution for $N$ is shown in Figure~\ref{fig:gliese581_N},
and shows strong evidence for at least eight periods
($P(N \leq 7 | \bdata) = 0.004$).
It is now recognised that
there are many possible sources for oscillations in a data set and not all
such oscillations should be claimed as planets \citep{2014Sci...345..440R}. 
Our model cannot distinguish between oscillations due to planets and
oscillations due to stellar activity: any oscillations found in the dataset
will be described as ``planets'' by the model. However it is interesting that
we find many more signals in the data than previous authors.

@@

The marginal likelihood was $\ln(\mathcal{Z}) \approx -618.04$ and the
information was $\mathcal{H} \approx 140.2$ nats. This compares favourably
to the marginal likelihood of $-640.1$ (for a 6-planet model)
found by \citet{fengji}, although it is unclear whether we used exactly the
same dataset.
Interestingly, the log-likelihood curve (Figure~\ref{fig:logl})
shows this problem has two phase transitions. While these do not affect the
posterior distribution (as they did for $\nu$ Oph), they would cause difficulties
if we tried to calculte the marginal likelihood using thermal annealing.

>>>>

% Bold symbols
%\renewcommand{	heta}{\boldsymbol{\theta}}
%\newcommand{D}{\boldsymbol{D}}
\newcommand{\bx}{\{x_i\}}

@@

\section{Inference}
Bayesian inference is the use of probability theory to describe uncertainty
\citep{sivia, ohagan}. In this framework, we approach data analysis problems by first
constructing a {\it hypothesis space}, which is the set of possible answers
to the problem we are considering. Normally, this is the set of possible
values of a vector of parameters $	heta$ whose values we want
to know. We then assign probability distributions
called the {\it prior} and the {\it sampling distribution}. The prior
distribution $p(	heta)$ describes
our initial uncertainty about which values of the parameters
$	heta$ are plausible, and the
sampling distribution $p(D | 	heta)$ describes
our initial uncertainty about the data set we're going to observe, as a
function of the unknown parameters $	heta$.
When the data is known, our state of knowledge about the parameter is
updated from the prior $p(\theta)$ to the posterior distribution given by
Bayes' rule:
\begin{eqnarray}
p(	heta | D) &=&
\frac{p(	heta)p(D | 	heta)}
{p(D)}
\end{eqnarray}
where $p(D | 	heta)$ is called the likelihood, once the actual dataset
is substituted in.
The denominator, often called the {\it evidence} or
{\it marginal likelihood}, is given by the expected value of the likelihood
with respect to the prior:
\begin{eqnarray}
\mathcal{Z} = p(D) = \int p(	heta) p(D | 	heta) d^n 	heta
\end{eqnarray}
where the integral is over the entire $n$-dimensional parameter space.

\subsection{Inferring the number of planets}
The number of orbiting planets, $N$, is an important parameter.
To calculate the posterior distribution for $N$, most authors
consider various trial values of $N$, and calculate the marginal likelihood
\begin{eqnarray}
p(D | N) &=& \int p(	heta | N) p(D | 	heta, N) \, d^n 	heta
\end{eqnarray}
for each possible value of $N$
\citep[e.g.][]{2011MNRAS.415.2523G, 2011MNRAS.415.3462F, 2014MNRAS.437.3540F, fengji}, marginalising
over all other model parameters.
The posterior distribution for $N$ can then be found straightforwardly by
using Bayes' rule with $N$ as the only unknown parameter:
\begin{eqnarray}
p(N | D) &=& \frac{p(N)p(D | N)}{\sum_N p(N)p(D | N)}.
\end{eqnarray}
Popular
methods for calculating the marginal likelihood are Nested Sampling
\citep{skilling} and ideas related to thermodynamic integration
\citep[e.g.][]{neal}.

@@

Imagine a high-dimensional unimodal posterior distribution that is composed of a
broad, high volume but low density ``slab'' with a narrow, low volume but high
density ``spike'' on top of it.
An example is a mixture of two concentric high-dimensional gaussians with
different widths. If you ran MCMC on such a posterior, it would be difficult
to jump between the slab and the spike components. If the MCMC is currently
in the spike region (or phase) it will be unable to escape: a proposed move
into the slab will be rejected because of the ratio of densities. Conversely,
if the MCMC was in the slab region, it would be unlikely to go into the spike
region, because its volume is so small: it would be very unlikely to
{\it propose} to move into the spike. Thus, the situation behaves much like
a multimodal posterior, despite only being unimodal.

Even if the slab contains a very small amount of posterior mass, its existence
can still cause problems with the calculation of the marginal likelihood
if annealing methods are used. The thermodynamic integral formula gives
the log of the marginal likelihood as an average of log likelihoods:
\begin{eqnarray}
\log(\mathcal{Z}) &=& \int_0^1 \left<\log\left[L(\theta)\right]\right>_\beta \, d\beta
\end{eqnarray}
where the expectation is taken with respect to the
distribution with ``inverse temperature'' $\beta$, proportional to
$\pi(\theta)L(\theta)^\beta$.

Nested Sampling \citep{skilling}, and variants such as DNS, are not affected
by phase transitions.

@@

\begin{table*}
\begin{tabular}{|l|l|l|}
\hline
Quantity	&	Meaning		& Prior\\
\hline
{\bf Hyperparameters}	&	\\
$N$		& Number of planets	& Uniform$(\{0, 1, ..., N_{\rm max}\})$\\
$\mu_T$		&	Median orbital period	& $\log(\mu_T) \sim$ Cauchy$(5.9, 1)T(-15.3, 27.1)$\\
$w_T$		&	Diversity of orbital periods & $w_T \sim$ Uniform$(0.1, 3)$\\
$\mu_A$		&	Mean amplitude	& $\log(\mu_A) \sim$ Cauchy$(0, 1)T(-21.2, 21.2)$\\
\hline
{\bf Planet Parameters}\\
$T_i$		&	Orbital period	&	$\log(T_i) \sim $ Biexponential($\log(\mu_T), w_T$)\\
$A_i$		&	Amplitude of signal	& Exponential$(\mu_A)$\\
$\phi_i$	&	Phase of signal	&	Uniform($0, 2\pi$)\\
$v_i$		&	Initial speed of orbit	&	Uniform$(0.4, 1)$\\
$\theta_i$	&	Viewing angle	&	Uniform$(0, 2\pi)$\\
\hline
{\bf Other}\\
$\sigma_{\rm extra}$	& ``Extra noise'' parameter	& $\log(\sigma) \sim$ Cauchy$(0, 1)T(-21.2, 21.2)$\\
$\nu$		& Shape parameter for $t$-distribution for noise & $\log(\nu) \sim$ Uniform$(\log(0.01), \log(1000))$\\
\hline
{\bf Data}\\
$Y_i$		& Radial velocity measurements	&
		Student-$t\left(m(t_i), \sqrt{\sigma_i^2 + \sigma_{\rm extra}^2}, \nu\right)$
\end{tabular}
\caption{All of the prior distributions in our Bayesian model.
The priors for the planet parameters are defined conditional on the values
of the hyperparameters. Uniform priors were used for parameters like phases.
For parameters where a rough initial guess is possible, heavy-tailed Cauchy
distributions were used so this information could be taken into account
in a non-dogmatic way. Time units are in days and amplitude units are
in metres per second. The maximum number of planets, $N_{\rm max}$, was
set to 15.\label{tab:priors}}
\end{table*}

@@

The posterior distribution for the logarithms of the periods is shown in
Figure~\ref{fig:nu_oph_periods}. Because of the label-switching degeneracy,
the posterior samples for all periods were combined to make this figure, which
shows three prominent peaks. Defining the log-periods by
$S_i =  \log_{10}\left[T/(\textnormal{1 day})\right]$,
Figure~\ref{fig:nu_oph_periods} is a Monte Carlo representation
of the mixture distribution
\begin{eqnarray}
f(S) &=& \sum_{N=0}^{10} p(N | D)\sum_{i=1}^N p(S_i | N, D).
\end{eqnarray}
The two rightmost peaks are the commonly accepted periods, and the third
prominent peak corresponds to a 75.7 $\pm$ 1.8 day period. The probability
this signal exists (technically, the probability a period exists between
67.6 and 85.1 days)
is 70\%, and given that it exists, its velocity amplitude
is 18.6 $\pm$ 8.3 m s$^{-1}$ and its eccentricity (or velocity parameter $v$)
is not well determined.

@@

Figure~\ref{fig:logl0} shows the relationship between the likelihood $L$
and the enclosed prior mass $X$ for the $\nu$ Oph analysis. These plots are
a standard output of Nested Sampling \citep{skilling} analyses, and provide
insights into the structure of the problem. Concave-up regions of this curve
indicate phase transitions which can cause severe problems for annealing-based
methods, and sometimes even for sampling the posterior distribution. In this
analysis, the models with a third signal exist to the left of the phase
transition at $\log(X) \approx -75$, and models without the signal exist to the
right of the phase transition. Mixing between these two phases is crucial for
accurately computing the posterior probability that the third signal exists.
However, if we had simply tried to sample the posterior distribution, it would
not mix easily between these two phases and would give a misleading posterior
probability for the existence of the third signal.

The marginal likelihood for our model was
$\log(\mathcal{Z}) \approx -217.04$ and the information
(Kullback-Leibler divergence from the prior to the posterior)
was $\mathcal{H} \approx 80.0$ nats.

@@

The posterior distribution for $N$ is shown in Figure~\ref{fig:gliese581_N},
and shows strong evidence for at least eight periods
($P(N \leq 7 | D) = 0.004$).
It is now recognised that
there are many possible sources for oscillations in a data set and not all
such oscillations should be claimed as planets \citep{2014Sci...345..440R}. 
Our model cannot distinguish between oscillations due to planets and
oscillations due to stellar activity: any oscillations found in the dataset
will be described as ``planets'' by the model. However it is interesting that
we find many more signals in the data than previous authors.

@@

The marginal likelihood was $\log(\mathcal{Z}) \approx -618.04$ and the
information was $\mathcal{H} \approx 140.2$ nats. This compares favourably
to the marginal likelihood of $-640.1$ (for a 6-planet model)
found by \citet{fengji}, although it is unclear whether we used exactly the
same dataset.
Interestingly, the log-likelihood curve (Figure~\ref{fig:logl})
shows this problem has two phase transitions. While these do not affect the
posterior distribution (as they did for $\nu$ Oph), they would cause difficulties
if we tried to calculte the marginal likelihood using thermal annealing.
